Approximation as Differentiation

### 2025-07-09 main/approx-as-tangents.tex
Theorem number for Amadio and Curien ref.

For the purposes of \GPS, we are not interested in using
approximations to model computation on infinite objects, but instead
for revealing how programs explore their inputs when producing parts
of their output. Therefore, we ignore completeness properties of the
partially ordered sets we consider\footnote{\bob{we might want to bring this back when considering general recursive programs. Are bidomains \cite{berry79} useful? they have two orderings. See also Laird's Bistable Biorders.}}.

@@

When giving a denotational semantics for sequential programming
languages Scott-continuous functions are too permissive. Famously,
\citet{plotkin77lcf}'s Parallel OR (\exref{parallel-or}, below) is
continuous but does not explore its input in a way consistent with a
sequential implementation. Continuous functions can explore their
input in a non-deterministic way as long as the result is
deterministic. This non-determinism results in functions that whose
output cannot be assigned a unique minimal input that caused
it. Therefore, continuous functions are incompatible with the central
idea in \GPS that we should be able to identify a \emph{minimal} part
of the input that leads to a part of the output.

\emph{Stability} is a property that can be required of monotone functions that was invented by \citet{berry79} in an attempt to capture sequentiality. This was not successful (see the $\mathrm{gustave}$ function in \exref{parallel-or}), but we will see now how it is closely related to problem of computing forward and backwards maps of approximations needed in \GPS. A textbook description of stable functions in the context of domain theory is given by \citet[Chapter 12]{amadio-curien}. We start with a property that is weaker than stability, but is easier to motivate in connection with derivatives of smooth functions.

@@

\begin{proof}
  \AGDA. See \citet{amadio-curien} (FIXME: theorem reference) for the
  case when the posets are also cpos. The crucial technical step,
  identified by \citet{berry79}, is that the ordering on conditionally
  multiplicative functions is not the extensional ordering
  ($f \sqsubseteq_{\mathrm{ext}} g$ iff
  $\forall x. f(x) \sqsubseteq g(x)$) but instead the stable ordering:
  $f \sqsubseteq_{\mathrm{st}} g$ iff $f \sqsubseteq_{\mathrm{ext}} g$ and
  $\forall x, x'.\,x \sqsubseteq x' \Rightarrow f(x) = f(x') \land
  g(x)$.
\end{proof}

>>>>

For the purposes of \GPS, we are interested in using
approximations not to model computation on infinite objects, but instead
for revealing how programs explore their inputs when producing parts
of their output. Therefore, we ignore completeness properties of the
partially ordered sets we consider\footnote{\bob{we might want to bring this back when considering general recursive programs. Are bidomains \cite{berry79} useful? they have two orderings. See also Laird's Bistable Biorders.}}.

@@

When giving a denotational semantics for sequential programming
languages, Scott-continuous functions are too permissive. Famously,
\citet{plotkin77lcf}'s Parallel OR (\exref{parallel-or}, below) is
continuous but does not explore its input in a way consistent with a
sequential implementation. Continuous functions can explore their
input in a non-deterministic way as long as the result is
deterministic. This non-determinism results in functions that whose
output cannot be assigned a unique minimal input that account for
it. Therefore, continuous functions are incompatible with the central
idea in \GPS that we should be able to identify a \emph{minimal} part
of the input that leads to a part of the output.

\emph{Stability} is a property that can be required of monotone functions, and was invented by \citet{berry79} in an attempt to capture sequentiality. This was unsuccessful (see the $\mathrm{gustave}$ function in \exref{parallel-or}), but we will see now how it is closely related to the problem of computing forward and backwards maps of approximations needed in \GPS. A textbook description of stable functions in the context of domain theory is given by \citet[Chapter 12]{amadio-curien}. We start with a property that is weaker than stability, but is easier to motivate in connection with derivatives of smooth functions.

@@

\begin{proof}
  \AGDA. See \citet[Theorem 12.2.8]{amadio-curien} for the
  case when the posets are also cpos. The crucial technical step,
  identified by \citet{berry79}, is that the ordering on conditionally
  multiplicative functions is not the extensional ordering
  ($f \sqsubseteq_{\mathrm{ext}} g$ iff
  $\forall x. f(x) \sqsubseteq g(x)$) but instead the stable ordering:
  $f \sqsubseteq_{\mathrm{st}} g$ iff $f \sqsubseteq_{\mathrm{ext}} g$ and
  $\forall x, x'.\,x \sqsubseteq x' \Rightarrow f(x) = f(x') \land
  g(x)$.
\end{proof}
